Concurrent Programming in Harmony

### 2021-01-02 book/paper.tex
Latest

There are two operations on condition variables: \texttt{wait}
\index{wait}%
and
\texttt{signal}.
\index{signal}%
\autoref{fig:hoare} presents the semantics by giving an implementation of
\texttt{wait} and \texttt{signal} as well as the code that is used to enter
and exit monitor methods.
\texttt{wait} increments the condition's counter, releases the monitor mutex,
blocks while trying to acquire the condition's semaphore, and after resuming
decrements the counter---in much the same way as we have seen for split binary
semaphores (SBS).
\texttt{signal} checks to see if the condition's count is non-zero, and if so
releases the condition's semaphore,
and then blocks by trying to acquire the mutex again.

\autoref{fig:hoaretest} presents a bounded buffer implemented using
Hoare monitors.  It is written in much the same way you would if using
the SBS technique (see \autoref{ex:bbsbs}).
There is no \texttt{release\_one} method.  Instead, one can conclude
that \texttt{put} causes the queue to be non-empty, and
\texttt{signal} will check if there are any threads waiting for
this event.  If so, \texttt{signal} will pass control to one such thread
and, unlike \texttt{release\_one}, re-enter the critical
section afterwards by acquiring the \textit{mutex}.

@@

As in Hoare monitors, there is a hidden mutex associated with each monitor,
and the mutex is automatically acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variables that a thread can wait on.
Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
\index{notify}%
When a thread $p$ invokes \texttt{notify}, it does not immediately pass 
control to a thread that is waiting on the corresponding condition (if there
is such a thread).  Instead, $p$ remains in the critical section
until it leaves the monitor explicitly.  At that point, any thread that
was notified will have a chance to enter the critical section, but they compete
with other threads trying to enter the critical section.
Basically, there is just one gate to enter the critical section, instead
of a main gate and a gate per waiting condition.

@@

This is a very important difference.  In Hoare monitors, when a thread
enters through a waiting gate, it can assume that the condition associated
with the waiting gate still holds because no other thread can run in between.
Not so with Mesa monitors: by the time a thread that was notified enters
through the main gate, other threads may have entered first and falsified
the condition.  So in Mesa, threads always have to check the condition
again after resuming from the \texttt{wait} operation.  This is accomplished
by wrapping each \texttt{wait} operation in a \textbf{while} statement that
loops until the condition of interest becomes valid.
A Mesa monitor therefore is more closely related to \emph{busy waiting}
than to split binary semaphores.

@@

The so-called
``Mesa monitor semantics'' or ``Mesa condition variable'' semantics
have become more popular than Hoare monitor semantics, and have been
adopted by all major programming languages.
That said, few programming languages provide full syntactical support
for monitors, instead opting to support monitor semantics through library
calls.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.
In addition, Java supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is entirely
through libraries.

@@

Like C, Java, and Python, Harmony does not have syntactical
support for monitors. Harmony provides support for monitors
through the Harmony \texttt{synch} module.
\autoref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition}() creates a new condition variable.
It is represented by a dictionary containing
a bag of contexts of threads waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

\texttt{wait} adds the context of the thread to the bag.
This increments the number of threads in the bag with the same context.
\texttt{wait} then waits until that count is restored to the
value that it had upon entry to \texttt{wait}.
\texttt{notify} removes an arbitrary context from
the bag, allowing one of the threads with that context to
resume and re-acquire the lock associated with the monitor.
\texttt{notifyAll} empties out the entire bag, allowing all threads
in the bag to resume.

@@

While Hoare moniters embody the split binary semaphore approach,
Mesa monitors are much closer to the busy-waiting approach.
The Mesa code then has to be careful to invoke \textbf{notify} or
\textbf{notifyAll} in the right places.
Much of the complexity of programming with Mesa condition variables
is in figuring out when to invoke \texttt{notify} and when to invoke
\texttt{notifyAll}.  As a rule of thumb: be conservative---it is
better to wake up too many threads than too few.  Waking up too
many threads may lead to some inefficiency, but waking up too few
may cause the application to get stuck.  Harmony can be particularly
helpful here, as it examines every corner case.  You can safely try
to replace each \texttt{notifyAll} with \texttt{notify} and see if
every possible execution of the application still terminates.

@@

Imagine five philosopers sitting around a table, each with a plate of food in
front of them and a fork between each two plates.  Each philosopher requires
two forks to eat.  To start eating, a philosopher first picks up the fork on
the left, then the fork on the right.  Each philosopher likes to take breaks
from eating to think for a while.  To do so, the philosopher puts down both
forks.  Each philosopher repeats this procedure.  Dijkstra had them repeating
this for ever, but for the purposes of this book, philosophers can leave
the table when they want to leave.

@@

Harmony allows modeling interrupts using the \texttt{trap} statement:
\begin{code}
\textbf{trap} \textit{handler} \textit{argument}
\end{code}
invokes \textit{handler argument} at some later, unspecified time.
Only one of these asynchronous events can be outstanding at a time; a new call to
\textbf{trap} overwrites any outstanding one.
\autoref{fig:trap} gives an example of how \textbf{trap} might be used.
Here, the \textit{main}() thread loops until the interrupt has occurred and
the \textit{done} flag has been set.

>>>>

There are two operations on condition variables: \texttt{wait}
\index{wait}%
and
\texttt{signal}.
\index{signal}%
\autoref{fig:hoare} presents the semantics by giving an implementation of
\texttt{wait} and \texttt{signal} as well as the code that is used to enter
and exit monitor methods.
\texttt{wait} increments the condition's counter, releases the monitor mutex,
blocks while trying to acquire the condition's semaphore, and after resuming
decrements the counter---in much the same way as we have seen for split binary
semaphores (SBS).
\texttt{signal} checks to see if the condition's count is non-zero, if so
releases the condition's semaphore,
and then blocks by trying to acquire the mutex again.

\autoref{fig:hoaretest} presents a bounded buffer implemented using
Hoare monitors.  It is written in much the same way you would if using
the SBS technique (see \autoref{ex:bbsbs}).
There is no \texttt{release\_one} method.  Instead, one can conclude
that \texttt{put} guarantees that the queue will be non-empty, and
\texttt{signal} will check if there are any threads waiting for
this event.  If so, \texttt{signal} will pass control to one such thread
and, unlike \texttt{release\_one}, re-enter the critical
section afterwards by acquiring the \textit{mutex}.

@@

As in Hoare monitors, there is a hidden mutex associated with each monitor,
and the mutex is automatically acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variables that a thread can wait on.
Like in Hoare monitors, the \texttt{wait} operation releases the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation \texttt{notify} rather than \texttt{signal}.
\index{notify}%
When a thread $p$ invokes \texttt{notify}, it does not immediately pass 
control to a thread that is waiting on the corresponding condition (if there
is such a thread).  Instead, $p$ remains in the critical section
until it leaves the monitor explicitly
(by calling either \texttt{release()} or \texttt{wait()}).
At that point, any thread that
was notified will have a chance to enter the critical section, but they compete
with other threads trying to enter the critical section.
Basically, there is just one gate to enter the critical section, instead
of a main gate and a gate per waiting condition.

@@

This is a very important difference.  In Hoare monitors, when a thread
enters through a waiting gate, it can assume that the condition associated
with the waiting gate still holds because no other thread can run in between.
Not so with Mesa monitors: by the time a thread that was notified enters
through the main gate, other threads may have entered first and falsified
the condition.  So in Mesa, threads always have to check the condition
again after resuming from the \texttt{wait} operation.  This is accomplished
by wrapping each \texttt{wait} operation in a \textbf{while} statement that
loops until the condition of interest becomes valid.
A Mesa monitor therefore is more closely related to busy waiting
than to split binary semaphores.

@@

The so-called
``Mesa monitor semantics'' or ``Mesa condition variable semantics''
have become more popular than Hoare monitor semantics and have been
adopted by all major programming languages.
That said, few programming languages provide full syntactical support
for monitors, instead opting to support monitor semantics through library
calls.
In Java, each object has a hidden lock \emph{and} a hidden condition variable
associated with it.
Methods declared with the \texttt{synchronized} keyword automatically
obtain the lock.  Java objects also support \texttt{wait}, \texttt{notify},
and \texttt{notifyAll}.
In addition, Java supports explicit allocations of locks
and condition variables.
In Python, locks and condition variables must be explicitly declared.
The \texttt{with} statement makes it easy to acquire and release a lock
for a section of code.
In C and C++, support for locks and condition variables is entirely
through libraries.

@@

Harmony provides support for monitors
through the Harmony \texttt{synch} module.
\autoref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition}() creates a new condition variable.
It is represented by a dictionary containing
a bag of contexts of threads waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

\texttt{wait} adds the context of the thread to the bag.
This increments the number of threads in the bag with the same context.
\texttt{wait} then loops until that count is restored to the
value that it had upon entry to \texttt{wait}.
\texttt{notify} removes an arbitrary context from
the bag, allowing one of the threads with that context to
resume and re-acquire the lock associated with the monitor.
\texttt{notifyAll} empties out the entire bag, allowing all threads
in the bag to resume.

@@

% While Hoare moniters embody the split binary semaphore approach,
% Mesa monitors are much closer to the busy-waiting approach.
When using Mesa condition variable you have to be careful to invoke
\textbf{notify} or \textbf{notifyAll} in the right places.
Much of the complexity of programming with Mesa condition variables
is in figuring out when to invoke \texttt{notify} and when to invoke
\texttt{notifyAll}.  As a rule of thumb: be conservative---it is
better to wake up too many threads than too few.
So in case of doubt, use \texttt{notifyAll}.
Waking up too
many threads may lead to some inefficiency, but waking up too few
may cause the application to get stuck.  Harmony can be particularly
helpful here, as it examines each and every corner case.
You can try to replace each \texttt{notifyAll} with \texttt{notify} and see
if every possible execution of the application still terminates.

@@

Imagine five philosopers sitting around a table, each with a plate of food in
front of them and a fork between every two plates.  Each philosopher requires
two forks to eat.  To start eating, a philosopher first picks up the fork on
the left, then the fork on the right.  Each philosopher likes to take breaks
from eating to think for a while.  To do so, the philosopher puts down both
forks.  Each philosopher repeats this procedure.  Dijkstra had them repeating
this for ever, but for the purposes of this book, philosophers can leave
the table when they are not using any forks.

@@

Harmony allows modeling interrupts using the \texttt{trap} statement:
\begin{code}
\textbf{trap} \textit{handler} \textit{argument}
\end{code}
invokes \textit{handler argument} at some later, unspecified time.
Thus you can think of \textbf{trap} as setting a timer.
Only one of these asynchronous events can be outstanding at a time;
a new call to \textbf{trap} overwrites any outstanding one.
\autoref{fig:trap} gives an example of how \textbf{trap} might be used.
Here, the \textit{main}() thread loops until the interrupt has occurred and
the \textit{done} flag has been set.
